---
title: Oceanic and Atmospheric Events with Severe Consequences on the Public Health
  and the Economy
author: "Alaettin Serhan Mete"
date: "5/25/2017"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

## Synopsis

This analysis involves exploring the U.S. National Oceanic and 
Atmospheric Administration's (NOAA) storm database to answer two 
basic but important questions: 
Which types of events are most harmful to population health, and
which types of events have the greatest economic consequences.
The analysis starts by obtaining the necessary dataset and 
getting it ready for analysis. Then, the actual analysis is performed
and the findings are summarized.

## Data Processing

As the initial step, let us load the pacakages that we'll use later on:

```{r}
# Load the dplyr package
if(!require(dplyr)){
    install.packages("dplyr")
    require(dplyr)
}

# Load ggplot2
if(!require(ggplot2)){
    install.packages("ggplot2")
    require(ggplot2)
}

# Load gridExtra for side-by-side plots
if(!require(gridExtra)) {
    install.packages("gridExtra")
    require(gridExtra)
}
```

Then download the necessary dataset and load it:
```{r cache = TRUE}
# Download if doesn't exist
if(!file.exists("StormData.csv.bz2")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  "StormData.csv.bz2",
                  method = "curl")
}

# Load into memory
input.dataset <- read.csv("StormData.csv.bz2")
```

Now let's convert this into a data-table that is more user friendly
and keep only the columns that'll be relevant to us:

```{r cache = TRUE}
# Convert into datatable
input.dataset.dt <- tbl_df(input.dataset)

# Let's select only the important columns
dataset <- select(input.dataset.dt,EVTYPE,FATALITIES:CROPDMGEXP)
```

## Results

Now let's analyze our dataset to answer our questions:

### Which types of events are most harmful to population health?

To answer this question, we'll compute the total number of fatalities 
and injuries for each event type. Then, we'll sort them in decreasing order:

```{r}
# Do the grouping and get the summary in one step
health.summary <- dataset %>% 
                    group_by(EVTYPE) %>%
                    summarise(total.deaths = sum(FATALITIES),
                              total.injuries = sum(INJURIES))
```

```{r}
# Order by deaths and print the leading 10 reasons
health.summary.death <- health.summary %>% arrange(desc(total.deaths))
head(health.summary.death, 10)
```

So, as the reader can see, the top three leading reasons of fatality are
tornadoes, excessive heat, and flash floods, respectively.

```{r}
# Order by injuries and print the leading 10 reasons
health.summary.injuries <- health.summary %>% arrange(desc(total.injuries))
head(health.summary.injuries, 10)
```

Again, as the reader can see, the top three leading reasons of injury are
tornadoes, tstm winds, and floods, respectively.

Now, let's construct a figure with two plots, the one on top showing the
total number of deaths (in log10) for the ten leading causes, 
and the one on the bottom showing the total number of injuries (again in log10) 
for the ten leading causes:

```{r}
#
# Deaths
#
# Do the ordering for the x-axis
death.sub <- health.summary.death[1:10,]
death.sub$EVTYPE <- as.character(death.sub$EVTYPE)
death.sub$EVTYPE <- factor(death.sub$EVTYPE, levels = unique(death.sub$EVTYPE))
# Construct the plot
p1 <- qplot(EVTYPE, log10(total.deaths), data = death.sub) 
p1 <- p1 + geom_line(lwd = 1, aes(group=1), color = "blue") + geom_point(size=2, shape=21, fill="blue")
p1 <- p1 + labs(title = "Total number of deaths for the ten leading event types (Log-scale)",
                x = "Event Type", y = expression(paste("Log"[10],"(Total deaths)")))
p1 <- p1 + theme(plot.title = element_text(hjust = 0.5))

#
# Injuries
#
# Do the ordering for the x-axis
injuries.sub <- health.summary.injuries[1:10,]
injuries.sub$EVTYPE <- as.character(injuries.sub$EVTYPE)
injuries.sub$EVTYPE <- factor(injuries.sub$EVTYPE, levels = unique(injuries.sub$EVTYPE))
# Construct the plot
p2 <- qplot(EVTYPE, log10(total.injuries), data = injuries.sub) 
p2 <- p2 + geom_line(lwd = 1, aes(group=1), color = "red") + geom_point(size=2, shape=21, fill="red")
p2 <- p2 + labs(title = "Total number of injuries for the ten leading event types (Log-scale)",
                x = "Event Type", y = expression(paste("Log"[10],"(Total injuries)")))
p2 <- p2 + theme(plot.title = element_text(hjust = 0.5))

# Now plot them on the same figure
grid.arrange(p1,p2,nrow = 2)
```

### Which types of events have the greatest economic consequences?

Now this bit is a bit more involved. There seems to be two types of damages:
property-related and crop-related. The associated variables in the dataset are
PROPDMG and PROPDMGEXP for the former, and CROPDMG and CROPDMGEXP for the latter.
After surfing the documents linked in the Coursera assignment page, as well as Google,
it's becomes apparent that the EXP variables hold the exponenet information with the 
allowed values of (the reader can find a bit more detailed discussion [here](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html)):

```{r}
unique(dataset$PROPDMGEXP)
```

For these, we'll assume they correspond to:

* **(blank), (?), (-)** : 0
* **(+)**  : 10^0
* **0..8** : 10^1
* **h,H**  : 10^2
* **k,K**  : 10^3
* **m,M**  : 10^6
* **b,B**  : 10^9

So, if the *PROPDMG = 20* and *PROPDMGEXP = k*, then the total property damage in USD 
will be *20x10^3 = 20000* USD, etc.

Now let us dechiper the dataset accordingly:

```{r cache = TRUE}
# Define a simple function to do the conversion
convert <- function(x) {
    value = toupper(as.character(x))
    switch(value,
           "+" = 1e0,
           "0" = 1e1,
           "1" = 1e1,
           "2" = 1e1,
           "3" = 1e1,
           "4" = 1e1,
           "5" = 1e1,
           "6" = 1e1,
           "7" = 1e1,
           "8" = 1e1,
           "H" = 1e2,
           "K" = 1e3,
           "M" = 1e6,
           "B" = 1e9,
           0)
}

# Now let's add damage values to the dataset
dataset.light <- dataset %>% 
                    rowwise() %>% 
                    mutate(propdmgusd = PROPDMG * convert(PROPDMGEXP),
                           cropdmgusd = CROPDMG * convert(CROPDMGEXP)) %>% 
                    select(EVTYPE,propdmgusd,cropdmgusd)
```

Now let's compute the total damages per event type:

```{r}
# Get the summary
dataset.light.summary <- dataset.light %>%
                            group_by(EVTYPE) %>%
                            summarise(totalpropdmg = sum(propdmgusd),
                                      totalcropdmg = sum(cropdmgusd))
```

and sort them by the total property and crop damages in descending order:

```{r}
# Order by total property damage and print the leading 10 reasons
dataset.light.summary.by.prop <- dataset.light.summary %>%
                                    arrange(desc(totalpropdmg))
head(dataset.light.summary.by.prop,10)
```

So, as the reader can see, the top three leading reasons causing property damage are
floods, hurricanes/typhoons, and tornadoes, respectively.

```{r}
# Order by total crop damage and print the leading 10 reasons
dataset.light.summary.by.crop <- dataset.light.summary %>%
                                    arrange(desc(totalcropdmg))
head(dataset.light.summary.by.crop,10)
```

As far as the crop damages are concerned, the top three leading reasons are
droughts, floods, and river floods (followed closely by ice storms), respectively.

Finally, let's make a figure with two plots:
total property damages (in millions USD) for the ten leading event types on the top, and
total crop damages (in millions USD) for the ten leading event types on the bottom:

```{r}
#
# Property
#
# Do the ordering for the x-axis
prop.sub <- dataset.light.summary.by.prop[1:10,]
prop.sub$EVTYPE <- as.character(prop.sub$EVTYPE)
prop.sub$EVTYPE <- factor(prop.sub$EVTYPE, levels = unique(prop.sub$EVTYPE))
# Construct the plot
p3 <- qplot(EVTYPE, totalpropdmg*1e-6, data = prop.sub) 
p3 <- p3 + geom_line(lwd = 1, aes(group=1), color = "blue") + geom_point(size=2, shape=21, fill="blue")
p3 <- p3 + labs(title = "Total property damage (in millions USD) for the ten leading event types",
                x = "Event Type", y = "Total property damage (in millions USD)")
p3 <- p3 + theme(plot.title = element_text(hjust = 0.5))

#
# Crop
#
# Do the ordering for the x-axis
crop.sub <- dataset.light.summary.by.crop[1:10,]
crop.sub$EVTYPE <- as.character(crop.sub$EVTYPE)
crop.sub$EVTYPE <- factor(crop.sub$EVTYPE, levels = unique(crop.sub$EVTYPE))
# Construct the plot
p4 <- qplot(EVTYPE, totalcropdmg*1e-6, data = crop.sub) 
p4 <- p4 + geom_line(lwd = 1, aes(group=1), color = "red") + geom_point(size=2, shape=21, fill="red")
p4 <- p4 + labs(title = "Total crop damage (in millions USD) for the ten leading event types",
                x = "Event Type", y = "Total crop damage (in millions USD)")
p4 <- p4 + theme(plot.title = element_text(hjust = 0.5))

# Now plot them on the same figure
grid.arrange(p3,p4,nrow = 2)
```







